Gender Segregation in Occupations: Preferences or Homophily?

### 2019-07-12 paper/Paper_2019_July_v1.tex
before yellowstone broad strokes reorganized the model and empirical sections, still needs rewriting

@@

\subsection{Payoff Functions}

Jobs choose the gender of worker that will maximize productivity per dollar spent, taking into account compensating differentials for job amenities, which may be differentially valued by men and women. This is consistent with a model of firms where jobs within a firm are filled independently, perhaps due to costly coordination or low complementarity. Individual jobs optimizing over rate of return does not imply a specific firm-level production function, but it produces similar behavior as firms minimizing cost with production quotas. Without data on firm size, I take this as an approximation of firm behavior.

@@

The ratio in equation \ref{firm} is used as opposed to an additive profit function for two reasons. First, since firms in the model only fill one vacancy at a time, rate of return is a reasonable object for firms to care about that does not require making assumptions about the number of total vacancies at a firm. By allowing the firm to maximize return on a single vacancy at a time I make a reasonable job-level approximation of a firm level profit function. Second, it is common in labor economics to specify wages as lognormally distributed. The multiplicative specification allows model wages to be lognormal in the estimation stage, which in turn provides the best possible fit to the observed wage data.

@@

The payoff to the firm of not hiring any worker, $\pi^N_j$, is normalized to one, so that the log payoff to not filling a vacancy is zero, ($ \bar{\pi}^N_j = 0$). There is no equivalent on the firm side to the idiosyncratic preference for non-employment that exists on the worker side. Heterogeneity in the outside option for a firm could reflect heterogeneity in the cost of hiring for different jobs, or the substitutability of a particular job in the firm production function.

\subsection{Job}
\begin{assumption}
Let the heterogeneity in job amenities, $\xi^g_j$, be distributed lognormal, and independently across $j$, such that $\bar{\xi}^g_j = ln(\xi^g_j)$ is distributed normally with location parameter normalized to zero and scale parameters for each gender, $\sigma^F_{\xi}$ and $\sigma^M_{\xi}$.
\end{assumption}
%
%Let $\Phi_{0,\sigma^g_{\xi}}$ and $\phi_{0,\sigma^g_{\xi}}$ are the cdf and pdf of the normal distribution with location zero and scale $\sigma^g_{\xi}$. The normality assumption implies the following choice probability.
%
%\begin{align*}
%Pr( \text{firm hires } F ) &=  Pr(\widebar{WTP}^F_o - \bar{W}^F_o + \xi^F_j \geq \widebar{WTP}^M_o - \bar{W}^M_o + \xi^M_j)\\
%&=  Pr(\widebar{WTP}^F_o - \bar{W}^F_o + \xi^F_j - (\widebar{WTP}^M_o - \bar{W}^M_o) \geq  \xi^M_j)\\
%&= \int_{-\infty}^{\infty} {\Phi_{0,\sigma^M_{\xi}}(\widebar{WTP}^F_o - \bar{W}^F_o - \widebar{WTP}^M_o - \bar{W}^M_o + \xi^F_j)}  \phi_{0,\sigma^F_{\xi}}(\xi^F_j) d\xi^F_j\\
%\end{align*}

I make the assumption of a fixed outside option because then the model is identified without use of data on vacancies, allowing me to test robustness to the use of the vacancy data. I can estimate the willingness-to-way of the firms with the maximum observed wage by gender and occupation. This strategy is less stable than using vacancy data. A firm will only hire a worker if the willingness-to-pay for that worker is higher than the wage the worker will accept in equilibrium, which means the payoff $ \pi^g_j $ must be greater than one.

The above assumption implies that jobs are unfilled when both $\bar{\xi}^M_j$ and $\bar{\xi}^F_j$ are very high. High $\bar{\xi}^M_j$ and $\bar{\xi}^F_j$ means very high dis-amenity value to both men and women, requiring high wages to compensate. So if a job is unattractive enough, it will not be filled.

%In a competitive equilibrium firms will produce at the production level that minimizes average cost, or equivalently maximizes willingness-to-pay over cost.

@@

Workers choose a job $j$ to maximize log utility $\bar{u}^{i}_j$. The common taste parameter for non-employment $\bar{u}^g_N$, where non-employment means either not in the labor force or unemployed, is normalized to zero. It is also assumed that in non-employment workers receive no wages or value from job amenities. The payoff to non-employment ($\bar{u}^i_N$) is therefore  $ \bar{\eta}^i_N $, the idiosyncratic taste for non-employment.\footnote{The log utility parameters for working, $\bar{u}^g_o$, include the disutility of working at all and are expected to be negative. However the underlying utility parameters $u^g_o = exp(\bar{u}^g_o)$ will always be positive leading to complementarity in wage and non-wage utility. Similarly, ${\xi^g_j}$ will be assumed to be log-normally distributed and therefore always positive, so the higher the dis-amenity value of the job the lower the utility. }

Furthermore worker utility will be decomposed into a common component and a dependence on the fraction female in the occupation, which is the endogenous amenity of interest. Let $F_o$ be the fraction female in occupation $o$.

@@

So $\alpha^g_o$ and  $\gamma^g$ will be the parameters of interest, but $\bar{u}^{g}_o $ can be recovered from these parameters and the fraction female. %% PUT IN FOOTNOTE why sigmas matter, but CUT THIS
Lastly, I estimate four variance parameters in total: variance of taste for occupation for men and women, and variance of job dis-amenities for men and women. These parameters are important because they govern the elasticity of labor supply to changes in wages or other non-wage amenities such as the gender ratio in the counterfactual, as well as the demand response to changes in equilibrium wage. Unlike \citeA{Chiappori2015}, who uses many markets to identify these parameters, I use the observed wage distribution as suggested in \citeA{Salanie2014a}. However I estimate using maximum likelihood in order to use the information from the wage distribution efficiently, and because the lognormal assumption means that the scale parameters cannot be separately estimated from all other parameters, as noted above. For all these reasons maximum likelihood is the most natural strategy for my application.

@@

\subsection{Worker}

@@

\section{Preference for Gender Ratio} \label{stage2}

@@

Previous literature only identifies the sum of wage and non-wage utility. Because a key model parameter is how non-wage utility varies with occupation fraction female, it is critical that I separately identify non-wage utility from utility from wages. The identification strategy in this paper builds off empirical applications of matching models to marriage markets (e.g. \citeA{Choo2006} and \citeA{Chiappori2015}), but differs in several key respects.

The first step towards separate identification is to assume that observed wages are the transfers between job and worker that clear the market. Although some transfers in the labor market may be non-wage, such as enhanced benefits, wages are a natural first-order approximation. Previous literature applying transferable utility to the marriage market has assumed a logit assumption for unobserved transfers because logit is most tractable. I model wages as lognormal in order to match the observed wage distribution, which is greater than zero and has a long right tail.\footnote{Unlike the logit case, the variance of the distribution of the maximum of lognormals depends on other model parameters to be estimated, so the variance needs to be jointly estimated unlike in \citeA{Salanie2014a} for example.} Second, I define an unmatched worker as a workers who is unemployed or out of the labor force, and an unmatched job as a posted vacancy that has remained unfilled for a period of time. Since vacancy data is potentially a poor proxy for what it means for a job to be unmatched, I try to limit the importance of the vacancy data in my estimation procedure by choosing a specification that in principle does not require it, by assuming non-negative profit on the job side.

For clarity of identification, I estimate the two sides of the market, workers and jobs, separately. The job side of the model is a maximum likelihood estimation exploiting the full variation of the observed wage distribution at the individual match level. The worker side, by contrast, is an instrumental variables regression at the occupation-year level, where instruments provide clean variation in the fraction female and reservation wage to trace out labor supply.\footnote{Combining the two estimation steps using joint GMM is theoretically feasible and would have the advantage that all data moments are used to identify all parameters. However the large number of parameters (914 parameters at once) and the need to pool all years of data (over 7 million observations) makes joint estimation computationally unattractive.}

@@

Since we observe log wages $\widebar{Wage}^g_j$, the expectation of observed wages suggests itself as an estimator for $\bar{W}^g_o$. Unfortunately there is a wrinkle to this strategy. There is a selection problem in that the only wages that are observed are the wages that maximize the job's choice over male, female, or not hiring any worker. The choice to not hire at all is especially problematic since it implies a truncation point that depends on $W^g_o$ which is unknown. Likewise, the selection effects depend on the scale of the unobserved job heterogeneity, $\sigma^g_\xi $, which is unknown. The problem therefore suggests joint likelihood estimation with selection and truncation, where the selection takes the form of a Tobit Type 5, to estimate $WTP^g_o$, $W^g_o$ and $\sigma^g_{\xi}$. See Appendix \ref{sec.likelihood} for the likelihood function in the notation of \citeA{Amemiya1985} and my notation.

\subsubsection{Identification}

@@

\subsection{Overview of Endogeneity and Instruments}

Therefore in my second estimation strategy I introduce Bartik-style instruments that exploit variation in the industry composition of occupations to isolate changes over time in the fraction female and wage by occupation that are caused by labor demand shifters. The use of instruments should isolate labor demand shocks, thus tracing out the labor supply curve conditional on fraction female and wage level.

@@

 \subsection{Fixed Effects Specification}

>>>>

The identification strategy in this paper builds off empirical applications of matching models to marriage markets (e.g. \citeA{Choo2006} and \citeA{Chiappori2015}), but differs in that I use data on transfers (wages) to separately identify worker and firm payoffs. In the marriage market, a lack of data on transfers means that only the sum of wage and non-wage utility is identified.\footnote{\citeA{Fox2008c} is able to estimate payoffs from the interaction of characteristics of both sides of the market but not full payoff functions.}. Because my research question is how non-wage utility varies with occupation fraction female, it is critical that I separately identify non-wage utility from utility from wages.

The first step towards separate identification is to assume that observed wages are the transfers between job and worker that clear the market. Although some transfers in the labor market may be non-wage, such as enhanced benefits, wages are a natural first-order approximation. Previous literature applying transferable utility to the marriage market has assumed a logit assumption for unobserved transfers because logit is most tractable. I model wages as lognormal in order to match the observed wage distribution, which is greater than zero and has a long right tail.\footnote{Unlike the logit case, the variance of the distribution of the maximum of lognormals depends on other model parameters to be estimated, so the variance needs to be jointly estimated unlike in \citeA{Salanie2014a} for example.} Second, I define an unmatched worker to be a worker who is unemployed or out of the labor force, and an unmatched job to be a posted vacancy that has remained unfilled for a period of time. Since vacancy data is potentially a poor proxy for what it means for a job to be unmatched, I try to limit the importance of the vacancy data in my estimation procedure by choosing a specification that in principle does not require it, by assuming non-negative profit on the job side.

For clarity of identification, I estimate the two sides of the market, workers and jobs, separately. The job side of the model is a maximum likelihood estimation exploiting the full variation of the observed wage distribution at the individual match level by year. The worker side, by contrast, is an instrumental variables regression at the occupation-year level, where instruments provide clean variation in the fraction female and reservation wage to trace out labor supply.\footnote{Combining the two estimation steps using joint GMM is theoretically feasible and would have the advantage that all data moments are used to identify all parameters. However the number of parameters (914 parameters) and the need to pool all years of data (7 million observations) makes joint estimation computationally unattractive.}

@@

\subsection{Payoff Functions}
\subsection{Job}

%Jobs choose the gender of worker that will maximize productivity per dollar spent, which may be differentially valued by men and women. This is consistent with a model of firms where jobs within a firm are filled independently, perhaps due to costly coordination or low complementarity. Individual jobs optimizing over rate of return does not imply a specific firm-level production function, but it produces similar behavior as firms minimizing cost with production quotas. Without data on firm size, I take this as an approximation of firm behavior.

@@

The ratio in equation \ref{firm} is used as opposed to an additive profit function for two reasons. First, since firms in the model only fill one vacancy at a time, rate of return is a reasonable object for firms to care about that does not require making assumptions about the number of total vacancies at a firm. By allowing the firm to maximize return on a single vacancy at a time I make a reasonable job-level approximation of a firm level profit function. Second, it is common in labor economics to specify wages as lognormally distributed. The multiplicative specification allows model wages to be normal in the estimation stage, when wages are assumed to be lognormal. The log payoff to the firm $j$ is:

@@

\subsection{Worker}
%In a competitive equilibrium firms will produce at the production level that minimizes average cost, or equivalently maximizes willingness-to-pay over cost.

@@

Workers choose a job $j$ to maximize log utility $\bar{u}^{i}_j$. The common taste parameter for non-employment $\bar{u}^g_N$, where non-employment means either not in the labor force or unemployed, is normalized to zero. It is also assumed that in non-employment workers receive no wages or value from job amenities. The payoff to non-employment ($\bar{u}^i_N$) is therefore  $ \bar{\eta}^i_N $, the idiosyncratic taste for non-employment.

%\footnote{The log utility parameters for working, $\bar{u}^g_o$, include the disutility of working at all and are expected to be negative. However the underlying utility parameters $u^g_o = exp(\bar{u}^g_o)$ will always be positive leading to complementarity in wage and non-wage utility. Similarly, ${\xi^g_j}$ will be assumed to be log-normally distributed and therefore always positive, so the higher the dis-amenity value of the job the lower the utility. }

Furthermore worker utility will be decomposed into a common component and a dependence on the fraction female in the occupation, which is the endogenous amenity of interest. Let $F_o$ be the fraction female in occupation $o$. To avoid simultaneity, the fraction female is defined as the fraction female of the occupation in period $t-1$, that is, the fraction female that would be observed by the worker as they make their occupation decisions.

@@

@@

%%%% CUT AND REWRITE
 %% PUT IN FOOTNOTE why sigmas matter, but CUT THIS
%Lastly, I estimate four variance parameters in total: variance of taste for occupation for men and women, and variance of job dis-amenities for men and women. These parameters are important because they govern the elasticity of labor supply to changes in wages or other non-wage amenities such as the gender ratio in the counterfactual, as well as the demand response to changes in equilibrium wage. Unlike \citeA{Chiappori2015}, who uses many markets to identify these parameters, I use the observed wage distribution as suggested in \citeA{Salanie2014a}. However I estimate using maximum likelihood in order to use the information from the wage distribution efficiently, and because the lognormal assumption means that the scale parameters cannot be separately estimated from all other parameters, as noted above. For all these reasons maximum likelihood is the most natural strategy for my application.

\subsection{Estimation}

Estimation then follows from distributional assumptions on the unobserved heterogeneity on each side of the market. The equilibrium solution allows separate estimation of each side.

\begin{assumption}
Let the heterogeneity in job amenities, $\xi^g_j$, be distributed lognormal, and independently across $j$, such that $\bar{\xi}^g_j = ln(\xi^g_j)$ is distributed normally with location parameter normalized to zero and scale parameters for each gender, $\sigma^F_{\xi}$ and $\sigma^M_{\xi}$.
\end{assumption}

Since we observe log wages $\widebar{Wage}^g_j$, the expectation of observed wages suggests itself as an estimator for $\bar{W}^g_o$. Unfortunately there is a wrinkle to this strategy. There is a selection problem in that the only wages that are observed are the wages that maximize the job's choice over male, female, or not hiring any worker. The choice to not hire at all is especially problematic since it implies a truncation point that depends on $W^g_o$ which is unknown. Likewise, the selection effects depend on the scale of the unobserved job heterogeneity, $\sigma^g_\xi $, which is unknown.\footnote{Estimating this parameter is important because is governs the elasticity of labor supply to changes in wages or other non-wage amenities such as the gender ratio in the counterfactual, as well as the demand response to changes in equilibrium wage. Unlike \citeA{Chiappori2015}, who uses many markets to identify these parameters, I use the observed wage distribution as suggested in \citeA{Salanie2014a}.} The problem therefore suggests joint likelihood estimation with selection and truncation, where the selection takes the form of a Tobit Type 5, to estimate $WTP^g_o$, $W^g_o$ and $\sigma^g_{\xi}$. See Appendix \ref{sec.likelihood} for the likelihood function in the notation of \citeA{Amemiya1985} and my notation.

%The payoff to the firm of not hiring any worker, $\pi^N_j$, is normalized to one, so that the log payoff to not filling a vacancy is zero, ($ \bar{\pi}^N_j = 0$). There is no equivalent on the firm side to the idiosyncratic preference for non-employment that exists on the worker side. Heterogeneity in the outside option for a firm could reflect heterogeneity in the cost of hiring for different jobs, or the substitutability of a particular job in the firm production function.

%
%Let $\Phi_{0,\sigma^g_{\xi}}$ and $\phi_{0,\sigma^g_{\xi}}$ are the cdf and pdf of the normal distribution with location zero and scale $\sigma^g_{\xi}$. The normality assumption implies the following choice probability.
%
%\begin{align*}
%Pr( \text{firm hires } F ) &=  Pr(\widebar{WTP}^F_o - \bar{W}^F_o + \xi^F_j \geq \widebar{WTP}^M_o - \bar{W}^M_o + \xi^M_j)\\
%&=  Pr(\widebar{WTP}^F_o - \bar{W}^F_o + \xi^F_j - (\widebar{WTP}^M_o - \bar{W}^M_o) \geq  \xi^M_j)\\
%&= \int_{-\infty}^{\infty} {\Phi_{0,\sigma^M_{\xi}}(\widebar{WTP}^F_o - \bar{W}^F_o - \widebar{WTP}^M_o - \bar{W}^M_o + \xi^F_j)}  \phi_{0,\sigma^F_{\xi}}(\xi^F_j) d\xi^F_j\\
%\end{align*}

A firm will only hire a worker if the willingness-to-pay for that worker is higher than the wage the worker will accept in equilibrium, which means the payoff $ \pi^g_j $ must be greater than one.

The above assumption implies that jobs are unfilled when both $\bar{\xi}^M_j$ and $\bar{\xi}^F_j$ are very high. High $\bar{\xi}^M_j$ and $\bar{\xi}^F_j$ means very high dis-amenity value to both men and women, requiring high wages to compensate. So if a job is unattractive enough, it will not be filled. I make the assumption of a fixed outside option because then the model is identified without use of data on vacancies, allowing me to test robustness to the use of the vacancy data. I can estimate the willingness-to-way of the firms with the maximum observed wage by gender and occupation. This strategy is less stable than using vacancy data.

@@

%\section{Preference for Gender Ratio} \label{stage2}

@@

Therefore I introduce Bartik-style instruments that exploit variation in the industry composition of occupations to isolate changes over time in the fraction female and wage by occupation that are caused by labor demand shifters. The use of instruments should isolate labor demand shocks, thus tracing out the labor supply curve conditional on fraction female and wage level.

% \subsection{Fixed Effects Specification}

@@

\subsubsection{Identification of Firm Parameters}

@@

\subsection{Identification of Worker Parameters}

@@
